On Optimality of Finite Element Integration

### 2016-01-27 TOMS_Paper/assembly.tex
Many improvementes, including structured expressions

\subsubsection{Identification and exploitation of structure}
\label{sec:se-rln}
Finite element expressions can be seen as composition of operations between tensors. Often, the optimal implementation strategy for these operations is to be determined out of two alternatives. For instance, consider $J^{-1} \nabla x \cdot J^{-1} \nabla x$, with $J^{-1}$ being the inverse Jacobian matrix for the change of coordinates and $x$ a generic two-dimensional vector. The tensor operation will reduce to the scalar expression $(a x^0_i + b x^1_i) (a x^0_i + b x^1_i) + ...$, in which $x^0_i$ and $x^1_i$ represent components of $x$ that depend on $L_i$. To minimize the operation count for expressions of this kind, we have two options.
\begin{Strategy}
\label{strategy:i}
Eliminating temporal sharing through generalized code motion.
\end{Strategy}
\begin{Strategy}
\label{strategy:ii}
Eliminating spatial sharing first -- through product expansion and factorization -- and temporal sharing afterwards, again through generalized code motion.
\end{Strategy}
In the current example, we observe that, depending on the size of $L_i$, applying Strategy~\ref{strategy:ii} could reduce the operation count since the expression would be recast as $x^0_i x^0_i a a + x^0_i x^1_i (ab + ab) + x^1_i x^1_i c c + ...$ and some hoistable sub-expressions would be exposed. On the other hand, Strategy~\ref{strategy:i} would have no effect as  $x$ only depends on a single loop, $L_i$. In general, the choice between the two strategies depends on multiple factors: the loop(s) size, the increase in operation count due to expansion (in Strategy~\ref{strategy:ii}), and the gain in code motion. A second application of Strategy~\ref{strategy:ii} was provided in Figure~\ref{code:multi_loopnest}. With the objective of characterizing when these two strategies suffice to reach quasi-optimality, we introduce a particular class of expressions and its relationship with multilinear loop nests.
\begin{Def}[Structured expression]
\label{def:struct-expr}
We say that ``an expression is structured along a loop nest $\Lambda$'' if and only if for eliminating the spatial sharing induced by any $s_{\Lambda}$ -- a generic symbol depending on at least one loop in $\Lambda$ -- it is sufficient to factorize all occurrences of $s_{\Lambda}$ in the expression.
\end{Def}
\begin{Prop}
\label{prop:multi-struct}
An expression along a multilinear loop nest is structured.
\end{Prop}
\begin{proof}
This is an obvious consequence of multilinearity.
\end{proof}
If $\Lambda$ were an arbitrary loop nest, a given symbol $s_{\Lambda}$ could appear everywhere (e.g., $n$ times in a summand and $m$ times in another summand with $n \neq m$, as argument of a higher level function, in the denominator of a division), thus posing the challenge of finding the factorization that maximizes temporal sharing. If $\Lambda$ is instead a finite element integration loop nest, thanks to Proposition~\ref{prop:multi-struct} we can build the space of f-decreasing transformations for quasi-optimality by just composing Strategy~\ref{strategy:i} and Strategy~\ref{strategy:ii}. The challenge for Algorithm~\ref{algo:sharing-elimination} will be to select the best f-decreasing transformations.

>>>>

\subsubsection{Identification and exploitation of structure}
\label{sec:se-rln}
Finite element expressions can be seen as composition of operations between tensors. Often, the optimal implementation strategy for these operations is to be determined out of two alternatives. For instance, consider $J^{-1} \nabla x \cdot J^{-1} \nabla x$, with $J^{-1}$ being the inverse Jacobian matrix for the change of coordinates and $x$ a generic two-dimensional vector. The tensor operation will reduce to the scalar expression $(a x^0_i + b x^1_i) (a x^0_i + b x^1_i) + ...$, in which $x^0_i$ and $x^1_i$ represent components of $x$ that depend on $L_i$. To minimize the operation count for expressions of this kind, we have two options.
\begin{Strategy}
\label{strategy:i}
Eliminating temporal sharing through generalized code motion.
\end{Strategy}
\begin{Strategy}
\label{strategy:ii}
Eliminating spatial sharing first -- through product expansion and factorization -- and temporal sharing afterwards, again through generalized code motion.
\end{Strategy}
In the current example, we observe that, depending on the size of $L_i$, applying Strategy~\ref{strategy:ii} could reduce the operation count since the expression would be recast as $x^0_i x^0_i a a + x^0_i x^1_i (ab + ab) + x^1_i x^1_i c c + ...$ and some hoistable sub-expressions would be exposed. On the other hand, Strategy~\ref{strategy:i} would have no effect as  $x$ only depends on a single loop, $L_i$. In general, the choice between the two strategies depends on multiple factors: the loop(s) size, the increase in operation count due to expansion (in Strategy~\ref{strategy:ii}), and the gain in code motion. A second application of Strategy~\ref{strategy:ii} was provided in Figure~\ref{code:multi_loopnest}. These examples motivate the introduction of a particular class of expressions, for which the two strategies have notable importance.
\begin{Def}[Structured expression]
\label{def:struct-expr}
We say that ``an expression is structured along a loop nest $\Lambda$'' if and only if for eliminating the spatial sharing induced by any $s_{\Lambda}$ -- a generic symbol depending on at least one loop in $\Lambda$ -- it is sufficient to factorize all occurrences of $s_{\Lambda}$ in the expression.
\end{Def}
\begin{Prop}
\label{prop:multi-struct}
An expression along a multilinear loop nest is structured.
\end{Prop}
\begin{proof}
This is an obvious consequence of multilinearity.
\end{proof}
If $\Lambda$ were an arbitrary loop nest, a given symbol $s_{\Lambda}$ could appear everywhere (e.g., $n$ times in a summand and $m$ times in another summand with $n \neq m$, as argument of a higher level function, in the denominator of a division), thus posing the challenge of finding the factorization that maximizes temporal sharing. If $\Lambda$ is instead a finite element integration loop nest, thanks to Proposition~\ref{prop:multi-struct} the space of f-decreasing transformations for quasi-optimality can be built by simply composing Strategy~\ref{strategy:i} and Strategy~\ref{strategy:ii}. The challenge for Algorithm~\ref{algo:sharing-elimination} will then be to select the best f-decreasing transformations.
